A Thee-Step Approach for Assessing Landscape Connectivity via Simulated Dispersal: African Wild Dog Case Study

### 2022-02-10 Chapter_1/04_Manuscript/Manuscript_2022-02-07.tex
Various updates

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable to dispersal.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route. Circuit theory presumes a complete
random walk and fails to incorporate directional persistence. While these
assumptions can be relaxed by explicitly simulating dispersal trajectories
across the landscape, a unified approach for such simulations is lacking.

Here, we present a simple three-step approach to simulate dispersal and to
assess connectivity using empirical GPS movement data and a set of habitat
covariates. In step one, we use integrated step-selection functions to fit a
mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, we apply the parameterized model to
simulate dispersal across the study area. In step three, we derive a set of
three complementary connectivity maps; a heatmap highlighting frequently
traversed areas, a betweenness map pinpointing dispersal corridors, and a map of
inter-patch connectivity indicating the presence and intensity of functional
links between habitat patches. As a case study, we apply the three-step
approach to GPS data collected on dispersing African wild dogs (\textit{Lycaon
pictus}) inhabiting the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA) in southern Africa.

Using step-selection functions we successfully parametrized a detailed dispersal
model that described dispersing individuals habitat and movement preferences, as
well as potential interactions among the two. The model substantially
outperformed a model that omitted such interactions and enabled us to simulate
80,000 dispersal trajectories across the study area. By explicitly simulating
individuals, this approach not only requires fewer unrealistic assumptions about
dispersal, but also permits the calculation of a rich suite of connectivity
metrics that together provide a comprehensive view on landscape connectivity.
The three derived connectivity maps revealed several dispersal hotspots and
corridors across the extent of our study area, yet each map highlighted a
different aspect of landscape connectivity, thus emphasizing their complementary
nature. Overall, our case study demonstrates that a simulation-based approach
offers a simple yet powerful alternative to traditional connectivity modeling
techniques. The approach is therefore useful for a variety of applications in
ecological, evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and to
colonize unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}). Unfortunately, both models rest on assumptions that appear
unsuitable for dispersers, thus calling for the development of alternative
approaches. One promising alternative is to assess landscape connectivity via
simulated dispersal trajectories generated from individual-based movement models
(IBMMs, \citealp{Diniz.2019}). However, IBMMs require a large number of
subjective modeling decisions, thus rendering among-system comparisons
difficult.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. Moreover, because both models require static permeability or
resistance surfaces as input, they are unable to reflect the temporal dimension
of dispersal and therefore fail to allow statements about the expected duration
for moving between habitat patches \citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This
strictly shifts the focus from a structural to a more functional view on
connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, IBMMs generate movement sequentially, i.e. they generate a series
of steps, so that the temporal dimension of dispersal becomes explicit and
allows modeling autocorrelation between successive steps \citep{Diniz.2019}.
Finally, simulations from IBMMs do not enforce movement or connections towards
preconceived endpoints but allow individuals to adjust their route  ``on the
go'', thereby preventing biases arising from misplaced endpoints. Despite these
advantages, a unifying approach to simulate dispersal and assess connectivity
using IBMMs is lacking. Considering the large number of subjective decisions
entailed by IBMMs, an approach that streamlines and standardizes the application
of dispersal simulations to assess connectivity will, however, be critical to
safeguard comparability among studies.

% Proposed Solution: Three-Step approach
\subsection{Proposed Solution: Three-Step Approach}
Here, we propose and exemplify a simple three-step approach for simulating
dispersal and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, we combine GPS movement data of dispersing individuals with habitat
covariates to fit a mechanistic movement model via integrated step-selection
functions (ISSFs, \citealp{Avgar.2016}). We chose to use ISSFs because the
framework not only allows inference on the study species' habitat kernel (i.e.
its habitat preferences), but also its movement kernel (i.e. its movement
preferences/capabilities) and potential interactions among the two
\citep{Avgar.2016, Fieberg.2021}. In step two, we use the parametrized movement
model to simulate dispersal across the study area. Comparable simulations have
already been applied to estimate steady-state utilization distributions of
resident individuals \citep{Potts.2013, Signer.2017} and to model landscape
connectivity, yet disregarding interdependencies between habitat and movement
kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we convert the
simulated trajectories into three complementary connectivity maps; (i) a heatmap
revealing frequently traversed areas (e.g. \citealp{Hauenstein.2019,
Zeller.2020}), (ii) a betweenness-map delineating dispersal corridors and
bottlenecks (e.g. \citealp{BastilleRousseau.2018}), (iii) and a map of
inter-patch connectivity, depicting the presence and intensity of functional
links between habitat patches, as well as the average dispersal duration
required to realize those connections (e.g. \citealp{Gustafson.1996,
Kanagaraj.2013}).

@@

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed approach using GPS movement data
collected on dispersing African wild dogs (\textit{Lycaon pictus}). The African
wild dog is a highly mobile species whose population persistence heavily relies
on the availability of large, natural or semi-natural landscapes and a
sufficient degree of connectivity among remaining subpopulations. Once common
throughout sub-Saharan Africa, this species has disappeared from much of its
historic range, largely due to human persecution, habitat fragmentation, and
disease outbreaks \citep{Woodroffe.2012}. Wild dogs typically disperse in
single-sex coalitions \citep{McNutt.1996, Behr.2020} and are capable of
dispersing several hundred kilometers \citep{DaviesMostert.2012, Masenga.2016,
Cozzi.2020}. Although previous studies have investigated connectivity for this
species using LCPA \citep{Hofmann.2021} and CT \citep{Brennan.2020}, a more
comprehensive and mechanistic understanding of dispersal and connectivity is
missing (but see \citealp{Creel.2020}). Nevertheless, with about 6,000
free-ranging wild dogs remaining in fragmented subpopulations
\citep{Woodroffe.2012}, reliable information on dispersal behavior and landscape
connectivity is essential for the conservation of this endangered carnivore. We
anticipated that a connectivity assessment based upon our three-step approach
would overcome several of the conceptual shortcomings of traditional
connectivity models, while providing a more detailed view on movement behavior
during dispersal its implications for landscape connectivity.

@@

\section{Methods}
\subsection{Case Study}
\subsubsection{GPS Data}
We applied the three step approach presented in \Cref{GraphicalAbstract} to GPS
movement data from 16 dispersing African wild dog coalitions (7 female and 9
male coalitions). This data has been collected between 2011 and 2019 from a
free-ranging wild dog population in northern Botswana. During dispersal, GPS
collars recorded a fix every 4 hours and regularly transmitted data over the
Iridium satellite system. To ensure comparable time intervals between GPS fixes,
we removed any fixes that were not successfully obtained at the desired 4-hour
schedule (allowing for a tolerance of \( \pm \) 15 minutes). To prepare the data
for step-selection analysis, we converted the fixes (n = 4'169) into steps,
where each step represented the straight-line movement between two consecutive
GPS fixes \citep{Turchin.1998}. We only considered steps with equal
step-durations (i.e. 4 hours) for further analysis. We will refer to these steps
as ``realized steps''. Additional details on the data collection and preparation
can be found in \cite{Cozzi.2020} and \cite{Hofmann.2021}.

@@

\subsubsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA,
\Cref{StudyArea}a and b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, thus hosting a rich diversity of landscapes, ranging from savannah
to grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydro-geographical feature and the world's largest
flood-pulsing inland delta. Large portions of the KAZA-TFCA are formally
protected in the form of national parks (NPs) or other protected areas, yet a
considerable portion of the landscape remains human-dominated (e.g. roads,
agricultural sites, and settlements).

@@

\subsubsection{Habitat Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textsf{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To render the seasonal dynamics of
water-cover for the extent of the Okavango Delta, we applied an algorithm that
enabled us to obtain weekly updated raster-layers for \textsf{water-cover} and
\textsf{distance-to-water} from MODIS satellite imagery \citep{Wolski.2017,
Hofmann.2021}. This algorithm is now implemented in the {\tt floodmapr} package
(available on GitHub; \url{https://github.com/DavidDHofmann/floodmapr}). To
ensure a consistent resolution across habitat covariates, we coarsened or
interpolated all layers to a resolution of 250 m x 250 m. A detailed description
of how we prepared each habitat covariate is provided in \cite{Hofmann.2021}.

We performed all data preparations, spatial computations, and statistical
analysis in {\tt R}, version 3.6.6 \citep{R.2020}. Some helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Step 1 - Movement Model}
We combined the collected GPS data with habitat covariates and used ISSFs
\citep{Avgar.2016} to parametrize a mechanistic movement model. More
specifically, we paired each realized step with a set of 24 randomly generated
alternative steps. A realized and its 24 random steps together formed a stratum
that received a unique identifier. As suggested by \cite{Avgar.2016}, we
generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted to realized steps (scale \(\theta\) = 6'308 and shape \(k\) = 0.37).
Note that our approach of sampling turning angles from a uniform distribution
does not imply that we assume uniform turning angles, as we will account for
directionality later in the model \citep{Avgar.2016, Fieberg.2021}.

Along each realized and random step, we extracted values from underlying habitat
covariate layers using the {\tt velox} package \citep{Hunziker.2021} and we
computed averages of each covariate along the steps. Besides extracting
\textit{habitat covariates}, we also computed movement metrics that we used as
\textit{movement covariates} in the ISSF models \citep{Avgar.2016,
Fieberg.2021}. Specifically, we computed the step length (\textsf{sl}), its
natural logarithm (\textsf{log(sl)}), and the cosine of the relative turning
angle (\textsf{cos(ta)}), which is a measure of directionality
\citep{Turchin.1998}, for each step. Because wild dog activity is low during the
hot midday hours \citep{Cozzi.2012}, we additionally created the variable
\textsf{LowActivity}, indicating whether a step was realized during periods of
low wild dog activity (09:00 to 17:00 local time) or high wild dog activity
(17:00 to 09:00 local time). To facilitate model convergence, we standardized
all continuous covariates to a mean of zero and a standard deviation of one.
Correlations among covariates were low (\(|r| < 0.6\); \citealp{Latham.2011}),
so we retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}), where \(w(x)\) depended on the step's associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's relative selection
strengths \citealp{Avgar.2017} towards these covariates (\(\beta_1, \beta_2,
..., \beta_n\)):

@@

\noindent To estimate relative selection strenghts (i.e. the
\(\beta\)-coefficients), we used mixed effects conditional logistic regression
analysis, implemented through the r-package {\tt glmmTMB} \citep{Brooks.2017}.
The implementation of conditional logistic regression has been proposed by
\cite{Muff.2020} and allows to model random slopes. The method requires to fix
the variance of the stratum specific intercept to a large value, so we fixed it
to an arbitrary high value of \(10 ^ 6\) and used disperser identity to model
random slopes for all covariates.

Our movement model was based on a habitat selection model that was previously
developed for dispersing wild dogs (hereafter referred to as \textit{base
model}, \citealp{Hofmann.2021}). In the base model, no interactions among
habitat covariates and movement covariates were considered, so we here expanded
the model and allowed for such interactions, acknowledging that movement
preferences during dispersal could depend on habitat conditions (details in
Appendix A1). To determine the most parsimonious movement model among model
candidates, we ran stepwise forward model selection based on Akaike's
Information Criterion (AIC, \citealp{Burnham.2002}). More specifically, we
started with the base model and iteratively increased model complexity by adding
all possible interactions between movement and habitat covariates. We validated
the predictive power of the most parsimonious model using k-fold
cross-validation for case-control studies as described in \cite{Fortin.2009}.
This validation proves a significant prediction in case the Spearman rank
correlation coefficient of predicted step-ranks and associated frequencies under
the movement model is significantly greater than under the assumption of random
preferences (details in Appendix A2).

\subsection{Step 2 - Dispersal Simulation}
We used the most parsimonious movement model to simulate individual dispersal
trajectories across the study area. The simulation of a dispersal trajectory
resembled an ``inverted'' ISSF and was set up as follows. (1) We defined a
source point and assumed a random initial orientation of the simulated animal.
(2) Starting from the source point, we generated 25 random steps by sampling
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from our fitted gamma distribution. (3) Along each random step, we extracted and
averaged values from the habitat covariate layers and we computed the movement
metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)}. To ensure
compatible scales with the fitted movement model, we standardized covariate
values using means and standard deviations from the empirical data. (4) We
applied the parametrized movement model to predict the selection score \(w(x)\)
for each step using \ref{EQ1} and we converted predicted scores into
probabilities using \ref{EQ2}. (5) We randomly sampled one of the generated
random steps based on assigned probabilities and determined the animal's new
position. We repeated steps (2) to (5) until 2,000 steps were realized and we
repeated the simulation until a total of 80'000 dispersal trajectories was
reached.

As source points for the simulations, we distributed 50,000 points at random
locations inside protected areas that were large enough to host an average size
wild dog home range (i.e. \(>\) 700 km\textsuperscript{2};
\citealp{Pomilia.2015}). We placed another 30,000 points randomly inside the
buffer zone, mimicking potential immigration into the study area (Figure S1).

To mitigate edge effects and to deal with random steps leaving the study area,
we followed \cite{Koen.2010} and artificially expanded all covariate layers by a
100 km wide buffer zone. Within the buffer zone, we randomized covariate values
by resampling values from the original covariate layers. Through this buffer
zone, simulated dispersers were able to leave and re-enter the main study area.
In cases where random steps crossed the outer border of this buffer zone, we
resampled steps until they fully lied within the buffer zone, essentially
forcing simulated individuals to remain within the expanded study area.

>>>>

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
Dispersal of individuals contributes to long-term population persistence, yet
requires a sufficient degree of landscape connectivity. To date, connectivity
has mainly been investigated using least-cost analysis and circuit theory, two
methods that make assumptions that are hardly applicable to dispersal. While
these assumptions can be relaxed by simulating dispersal trajectories across the
landscape, a unified approach for such simulations is lacking.

% Least-cost analysis assumes that animals move towards a known endpoint and are
% knowledgeable about the most favorable route. Circuit theory presumes a
% complete random walk and fails to incorporate directional persistence.

Here, we present a simple three-step approach to simulate dispersal movements
and to assess connectivity using empirical GPS movement data and a set of
habitat covariates. In step one, integrated step-selection functions are used to
fit a mechanistic movement model describing habitat and movement preferences of
dispersing individuals. In step two, the parameterized model is used to simulate
dispersal trajectories across the study area. In step three, a set of three
complementary connectivity maps is derived: a heatmap that highlights frequently
traversed areas, a betweenness map that pinpoints dispersal corridors, and a map
of inter-patch connectivity that indicates the presence and intensity of
functional links between habitat patches. As a case study, we applied the
approach to GPS data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon pictus}) inhabiting the Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA) in southern Africa.

Besides providing insights into the movement and habitat preferences of
dispersing individuals, our movement model also rendered how movement behavior
depended on habitat conditions. The model thereby substantially outperformed a
model that omitted such interactions. While we simulated a total of 80,000
dispersal trajectories, metrics of connectivity only marginally changed beyond
10,500 simulations, suggesting that fewer simulations would have sufficed. The
three connectivity maps derived from simulations revealed several dispersal
hotspots and corridors across the extent of our study area. Each map emphasized
a different aspect of landscape connectivity, thus highlighting their
complementary nature. Overall, our case study demonstrates that a
simulation-based approach utilizing step-selection functions offers a simple yet
powerful alternative to traditional connectivity modeling techniques. Our
approach not only makes fewer unrealistic assumptions about dispersal but also
permits a more mechanistic understanding of dispersal movements and landscape
connectivity. It is thus useful for a variety of applications in ecological,
evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the ability
to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. Identifying dispersal corridors
not only necessitates a comprehensive understanding of the factors that limit
dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the most commonly
used models are least-cost path analysis (LCPA; \citealp{Adriaensen.2003}) and
circuit theory (CT; \citealp{McRae.2006, McRae.2008}). Unfortunately, both
models rest on assumptions that appear unsuitable for dispersers, thus calling
for the development of alternative approaches. One promising alternative is to
assess landscape connectivity via simulated dispersal trajectories generated
from individual-based movement models (IBMMs, \citealp{Diniz.2019}). However,
the large number of subjective modeling decisions required by IBMMs renders
among-system comparisons difficult.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be justifiable for
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory towards an unknown endpoint
\citep{Koen.2014, Cozzi.2020}. CT, on the contrary, posits that animals move
according to a random walk, entailing that autocorrelation between subsequent
movements cannot be rendered \citep{Diniz.2019}. For dispersers, however,
autocorrelated movements are regularly observed \citep{Cozzi.2020,
Hofmann.2021}, meaning that dispersal trajectories are usually strongly
directional. Moreover, because both models require static permeability or
resistance surfaces as input, they are unable to reflect the temporal dimension
of dispersal, such that statements about the expected duration for moving
between habitat patches are impossible \citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
using IBMMs and by converting simulated trajectories into meaningful measures of
connectivity \citep{Diniz.2019}. In contrast to LCPA and CT, IBMMs allow to
explicitly simulate how individuals move across and interact with the
encountered landscape \citep{Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020}, as well as to render potential interactions
between movement behavior and habitat conditions \citep{Avgar.2016}. This
strictly shifts the focus from a structural to a more functional view on
connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, IBMMs generate movement sequentially, i.e. they generate a series
of steps, so that the temporal dimension of dispersal becomes explicit and
allows modeling autocorrelation between consecutive steps \citep{Diniz.2019}.
Finally, simulations from IBMMs do not enforce movement or connections towards
preconceived endpoints, thereby preventing biases arising from misplaced
endpoints. Despite these advantages, a unifying framework to simulate dispersal
and assess connectivity using IBMMs is lacking. Considering the large number of
subjective decisions entailed by IBMMs, an approach that streamlines and unifies
the application of dispersal simulations to assess connectivity will, however,
be critical to safeguard interspecfic comparability and reproducability among
studies.

% Proposed Solution: Three-Step approach
\subsection{Proposed Solution: Three-Step Approach}
Here, we propose and exemplify a simple three-step approach for simulating
dispersal and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, GPS movement data of dispersing individuals is combined with habitat
covariates to fit a mechanistic movement model via integrated step-selection
functions (ISSFs, \citealp{Avgar.2016}). ISSFs allow inference on the study
species' habitat kernel (i.e. habitat preferences), its movement kernel (i.e.
movement preferences/capabilities), as well as potential interactions between
the two \citep{Avgar.2016, Fieberg.2021}. In step two, the parametrized movement
model is used to simulate individual dispersal trajectories starting from
randomly placed source points. Comparable simulations have already been applied
to estimate steady-state utilization distributions of resident individuals
\citep{Potts.2013, Signer.2017} and to model landscape connectivity, yet
disregarding interdependencies between habitat and movement kernels
\citep{Clark.2015, Zeller.2020}. Finally, in step three, the simulated
trajectories are converted into three complementary connectivity maps; (i) a
heatmap revealing areas that are frequently traversed by dispersers (e.g.
\citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map delineating
dispersal corridors and bottlenecks (e.g. \citealp{BastilleRousseau.2018}),
(iii) and a map of inter-patch connectivity, depicting the presence and
intensity of functional links between habitat patches, as well as the average
dispersal duration required for the realization of those connections (e.g.
\citealp{Gustafson.1996, Kanagaraj.2013}).

@@

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed approach (\Cref{GraphicalAbstract})
using GPS movement data collected on dispersing individuals of the endangered
African wild dog (\textit{Lycaon pictus}). The African wild dog is a highly
mobile species whose population persistence heavily relies on the availability
of large, natural or semi-natural landscapes and a sufficient degree of
connectivity among remaining subpopulations. Once common throughout sub-Saharan
Africa, this species has disappeared from much of its historic range, largely
due to human persecution, habitat fragmentation, and disease outbreaks
\citep{Woodroffe.2012}. Wild dogs typically disperse in single-sex coalitions
\citep{McNutt.1996, Behr.2020} and are capable of dispersing several hundred
kilometers \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}. Although
previous studies have investigated connectivity for this species using LCPA
\citep{Hofmann.2021} or CT \citep{Brennan.2020}, a more comprehensive and
mechanistic understanding of dispersal and connectivity is missing (but see
\citealp{Creel.2020}). Nevertheless, with about 6,000 free-ranging wild dogs
remaining in fragmented subpopulations \citep{Woodroffe.2012}, reliable
information on dispersal behavior and landscape connectivity is essential for
the conservation of this endangered carnivore. We anticipated that a
connectivity assessment based on our three-step approach would overcome several
of the above highlighted conceptual shortcomings of traditional connectivity
models, while providing a more detailed view on movement behavior during
dispersal.

@@

\section{Methods}
\subsection{Case Study}
\subsubsection{GPS Data}
We applied the three step approach mentioned above to GPS movement data from 16
dispersing African wild dog coalitions (7 female and 9 male coalitions). This
data has been collected between 2011 and 2019 from a free-ranging wild dog
population in northern Botswana (details on the data collection can be found in
\cite{Cozzi.2020} and \cite{Hofmann.2021}). During dispersal, GPS collars
recorded a fix every 4 hours and regularly transmitted data over the Iridium
satellite system. To ensure regular time intervals between GPS fixes, we removed
any fixes that were not successfully obtained at the desired 4-hour schedule
(allowing for a tolerance of \( \pm \) 15 minutes). To prepare the data for
step-selection analysis, we converted the fixes (n = 4'169) into steps, where
each step represented the straight-line movement between two consecutive GPS
fixes \citep{Turchin.1998}.

@@

\subsubsection{Study Area}
Our simulation of dispersal trajectories and assessment of connectivity spanned
across the entire Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA,
\Cref{StudyArea}a and b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, thus hosting a rich diversity of landscapes, ranging from savannah
to grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydro-geographical feature and the world's largest
flood-pulsing inland delta. Large portions of the KAZA-TFCA are formally
protected in the form of national parks (NPs) or other protected areas, yet a
considerable portion of the landscape is still human-dominated (e.g. roads,
agricultural sites, and settlements). We used NPs as patches to calculate
inter-patch connectivity (cf. Section 2.4.3). The decision to only consider NPs
as patches was purely out of simplicity and does not imply that dispersal
between other protected areas does not occur.

@@

\subsubsection{Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textsf{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To render the seasonal dynamics in
water-cover of major water bodies for the extent of the Okavango Delta, we
applied an algorithm that enabled us to obtain weekly updated raster-layers for
\textsf{water-cover} and \textsf{distance-to-water} from MODIS satellite imagery
\citep{Wolski.2017, Hofmann.2021}. This algorithm is now implemented in the {\tt
floodmapr} package (available on GitHub;
\url{https://github.com/DavidDHofmann/floodmapr}). To ensure a consistent
resolution across habitat covariates, we coarsened or interpolated all layers to
a resolution of 250 m x 250 m. A detailed description of how we prepared each
habitat covariate is provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in the ISSF models \citep{Avgar.2016, Fieberg.2021}.
Specifically, we computed for each step the step length (\textsf{sl}), its
natural logarithm (\textsf{log(sl)}), and the cosine of the relative turning
angle (\textsf{cos(ta)}), which is a measure of directionality
\citep{Turchin.1998}. Because wild dog activity is low during the hot midday
hours \citep{Cozzi.2012}, we additionally created the variable
\textsf{LowActivity}, indicating whether a step was realized during periods of
low wild dog activity (09:00 to 17:00 local time) or high wild dog activity
(17:00 to 09:00 local time). We performed all data preparations, spatial
computations, and statistical analysis in {\tt R}, version 3.6.6 \citep{R.2020}.
Some helper functions were written in {\tt C++} and imported into {\tt R} using
the {\tt Rcpp} package \citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Step 1 - Movement Model}
We used ISSFs \citep{Avgar.2016} to parametrize a mechanistic movement model
starting from GPS movement data collected on dispersing individuals. More
specifically, we paired each realized (i.e. observed) step with a set of 24
randomly generated alternative steps. A realized step plus its 24 random steps
formed a 25-step-stratum and received a unique identifier. As suggested by
\cite{Avgar.2016}, we generated random steps by sampling random turning angles
from a uniform distribution (\(-\pi, +\pi\)) and step lengths from a gamma
distribution that was fitted to realized steps (scale \(\theta\) = 6'308 and
shape \(k\) = 0.37).

Along each realized and random step, we extracted values from underlying habitat
covariate layers using the {\tt velox} package \citep{Hunziker.2021} and we
computed averages of each covariate along the steps. For each step, we further
calculated the movement metrics \textsf{sl}, \textsf{log(sl)}, and
\textsf{cos(ta)}. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one.
Correlations among covariates were low (\(|r| < 0.6\); \citealp{Latham.2011}),
so we retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) to each step (\ref{EQ1};
\citealp{Fortin.2005}), where \(w(x)\) depended on the step's associated
covariates (\(x_1, x_2, ..., x_n\)) and on the animal's preferences (i.e.
relative selection strengths; \citealp{Avgar.2017}) towards these covariates
(\(\beta_1, \beta_2, ..., \beta_n\)):

@@

\noindent To estimate preferences (i.e. the \(\beta\)-coefficients), we used
mixed effects conditional logistic regression analysis \citep{Muff.2020},
implemented through the r-package {\tt glmmTMB} \citep{Brooks.2017}. The method
introduced by \cite{Muff.2020} allows to model random slopes, yet requires to
fix the variance of the stratum specific intercept to a large value. We
therefore fixed the variance of the stratum specific intercept to an arbitrary
high value of \(10 ^ 6\) and used disperser identity to model random slopes for
all covariates.

The structure of our movement model was based on a habitat selection model that
was previously developed for dispersing wild dogs (hereafter referred to as
\textit{base model}, \citealp{Hofmann.2021}). In the base model, no interactions
among habitat covariates and movement covariates were considered, so we here
expanded the model and allowed for interactions between all movement covariates
and habitat covariates, thus reflecting that movement behavior could depend on
habitat conditions (details in Appendix A1). To determine the most parsimonious
movement model among model candidates, we ran stepwise forward model selection
based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}). We
validated the predictive power of the most parsimonious model using k-fold
cross-validation for case-control studies as described in \cite{Fortin.2009}.
This validation proves a significant prediction in case the Spearman rank
correlation coefficient of predicted step-ranks and associated frequencies under
the movement model is significantly greater than under the assumption of random
preferences (details in Appendix A2).

\subsection{Step 2 - Dispersal Simulation}
We used the most parsimonious movement model to simulate 80'000 individual
dispersal trajectories across the study area. The simulation of a dispersal
trajectory resembled an ``inverted'' ISSF and was set up as follows. (1) We
defined a source point and assumed a random initial orientation of the simulated
animal. (2) Starting from the source point, we generated 25 random steps by
sampling turning angles from a uniform distribution (\(-\pi, +\pi\)) and step
lengths from our fitted gamma distribution. (3) Along each random step, we
extracted and averaged values from the different habitat covariate layers and
computed the movement metrics \textsf{sl}, \textsf{log(sl)}, and
\textsf{cos(ta)}. To ensure compatible scales with the fitted movement model, we
standardized covariate values using means and standard deviations from the
empirical data. (4) We applied the parametrized movement model to predict the
selection score \(w(x)\) for each step using \ref{EQ1} and we converted
predicted scores into probabilities using \ref{EQ2}. (5) We randomly sampled one
of the generated random steps based on assigned probabilities and determined the
animal's new position. We repeated steps (2) to (5) until 2,000 steps were
realized.

As source points for the simulations, we distributed 50,000 points at random
locations inside protected areas that were larger than the average home range of
resident wild dog packs (i.e. \(>\) 700 km\textsuperscript{2};
\citealp{Pomilia.2015}). We placed another 30,000 points randomly inside the
buffer zone, mimicking potential immigration into the study area (Figure S1).

To mitigate edge effects and to deal with random steps leaving the study area,
we followed \cite{Koen.2010} and artificially expanded all covariate layers by a
100 km wide buffer zone. Within the buffer zone, we randomized covariate values
by resampling values from the original covariate layers. Through this buffer
zone, simulated dispersers were able to leave and re-enter the main study area.
In cases where random steps crossed the outer border of this buffer zone, we
resampled steps until they fully lied within the buffer zone, essentially
forcing individuals to remain within the expanded study area.

To ensure reliable connectivity estimates, we determined the number of simulated
dispersal trajectories required for connectivity to reach a ``steady state''.
For this purpose, we distributed 1,000 rectangular ``checkpoints'', each with an
arbitrary extent of 5 km x 5 km, at random coordinates within the study area
(excluding the buffer). We then determined the relative frequency at which each
checkpoint was traversed by simulated dispersal trajectories (hereafter referred
to as relative traversal frequency) as we gradually increased the number of
simulated trajectories from 1 to 50,000. To assess variability in the relative
traversal frequency, we repeatedly subsampled 100 times from all 50'000
trajectories and computed the mean traversal frequency across replicates, as
well as its 95\% prediction-interval for each checkpoint. We considered
connectivity to have reached a steady state once the width of the
prediction-interval dropped below a value of 0.01 for all checkpoints.
